---
title: "The likert Package"
author: "Jason Bryer, Ph.D."
date: "2016-12-15"
output:
  ioslides_presentation:
    self_contained: true
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits=3)
options(width=120)
library(grid)
```

## Likert Scales

Invented by Rensis Likert (1932).

Questionaire items where respondents respond to a question on a sequence of ordered responses. Common response types include:

* Agreement (e.g. Strongly Disagree, Disagree, Neither agree nor disagree, Agree, Strongly Agree)
* Frequency (e.g. Almost Never, Not Very Often, Somewhat Often, Pretty Often, Almost Often)
* Quality (e.g. Very Good, Good, Acceptable, Poor, Very Poor )
* Relevance (e.g. Excellent, Somewhat, Poor)
* Value (e.g. High, Moderate, Low, None)
* Importance (e.g. Very Important, Important, Moderately Important, Slightly Important, Not Important)

Likert, R (1932). [A technique for the measurement of attitudes](http://www.voteview.com/pdf/Likert_1932.pdf). *Archives of Psychology. 140*: 1–55.

## Categorical vs. Continuous

A key question to ask when using Likert items: Is the variable categorical or contiuous?

When considering the response to be continuous you are assuming the intervals are equal? That is:

$$ Strongly Agree - Agree = Disagree - Neutral $$

## Motivation

This package aims to provide the necessary tools to allow the user to make his/her own decisions about how the data should be treated, while also providing figures that may help clarify how well the assumptions discussed here hold up.

* summarize likert type items
* provide tabular results
* provide visualizations
* allow flexibility (e.g. ggplot2 annotate)
* have a consistent aesthetic
* use familiar S3 generics (i.e. print, plot, summary)

## Example: PISA Reading Attitudes

The Programme of International Student Assessment (PISA; 2009) assess ~15 year-old students about their reading attitudes:

* I read only if I have to.
* Reading is one of my favorite hobbies.
* I like talking about books with other people.
* I find it hard to finish books.
* I feel happy if I receive a book as a present.
* For me, reading is a waste of time.
* I enjoy going to a bookstore or a library.
* I read only to get information that I need.
* I cannot sit still and read for more than a few minutes. 
* I like to express my opinions about books I have read. 
* I like to exchange books with my friends.

Students respond to one of four options: Strongly disagree, Disagree, Agree, Strongly Agree

## Example: PISA Reading Attitudes

```{r, message=FALSE}
library(likert)
```

Demo included in the package:

```{r, eval=FALSE}
demo('likert', package='likert')
```

Data preparation: 

```{r, message=FALSE}
data(pisaitems)
items24 <- pisaitems[,substr(names(pisaitems), 1,5) == 'ST24Q']
names(items24) <- c(ST24Q01 = "I read only if I have to.",
			        ST24Q02 = "Reading is one of my favorite hobbies.",
			        ST24Q03 = "I like talking about books with other people.",
			        ST24Q04 = "I find it hard to finish books.",
			        ST24Q05 = "I feel happy if I receive a book as a present.",
			        ST24Q06 = "For me, reading is a waste of time.",
			        ST24Q07 = "I enjoy going to a bookstore or a library.",
			        ST24Q08 = "I read only to get information that I need.",
			        ST24Q09 = "I cannot sit still and read for more than a few minutes.",
			        ST24Q10 = "I like to express my opinions about books I have read.",
			        ST24Q11 = "I like to exchange books with my friends.")
```

## Example: PISA Reading Attitudes

```{r}
str(items24)
```

## Using the `likert` function

```{r}
l24 <- likert(items24)
l24 #print(l24)
```

## Numeric Results

```{r}
summary(l24)
```

## Plotting Likert Results {.flexbox .vcenter}

```{r}
plot(l24)
```

## Plotting Likert Results (center) {.flexbox .vcenter}

The `center` option will define what category is considered to be the center.

```{r}
plot(l24, center=2)
```

## Plotting Likert Results (ordering) {.flexbox .vcenter}

```{r}
plot(l24, ordered=FALSE, group.order=names(items24)) #Specify the exact order of the y-axis
```

## Plotting Likert Results (changing colors) {.flexbox .vcenter}

```{r}
plot(l24, colors=c('orange','darkorange','darkblue','blue'))
```

## Plotting Likert Results (response histograms) {.flexbox .vcenter}

```{r}
plot(l24, include.histogram=TRUE)
```

## Density Plots {.flexbox .vcenter}

```{r}
plot(l24, type='density', facet=FALSE)
```

## Heatmap Plots {.flexbox .vcenter}

```{r}
plot(l24, type='heat', wrap=30, text.size=4)
```

## Grouping Responses

```{r}
l24g <- likert(items24[,1:5], grouping=pisaitems$CNT)
print(l24g)
```


## Grouping Responses: Summary

```{r}
summary(l24g)
```

## Grouping Responses: Plotting {.flexbox .vcenter}

```{r}
plot(l24g)
```

## Example 2: DAACS Pre-Pilot

**[daacs.net/shiny/PrePilot](http://daacs.net/shiny/PrePilot)**


